Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-06-16 paper.tex

%\paragraph{Evaluation}
After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction loss on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction loss (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1} = \finitepast\finitefuture$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)},
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $\finitepast$ as in (\ref{eq:z-gauss}), and prediction loss 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] \approx - \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(\finitefuture | Z).
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction loss achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$ \rljf{How?}. Because we consider stationary processes, we have 
$- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture)\right] = - \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$.
Using this observation, we estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~\cite{mcallester2018formal}:
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture) - \log P_\psi(\finitefuture | Z)\right],
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

During optimization, we add the cross-entropy term $- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$ to the PRD objective~(\ref{eq:bound}), so that the full training objective comes out to:
\begin{equation}\label{eq:bound-augmented}
	\min_{\phi, \psi, q,\eta} \left[\mathbb{E}_{\finitepast, \finitefuture} \left[ -	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} \left[ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] - \log P_\eta(\finitepast) \right]\right].
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
Again by Gibbs' inequality and Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}, this is minimized when $P_\eta$ represents the true distribution over length-$M$ blocks $P(\finitepast)$, $P_\phi(Z|\finitepast)$ describes an optimal code for the given $\lambda$, $q$ is its marginal distribution, and $P_\psi(\finitefuture|Z)$ is the Bayes-optimal decoder.
For approximate solutions to this augmented objective, the inequalities~(\ref{eq:bound1-new}-\ref{eq:bound2-new}) will also remain true due to Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}.

@@

Training used Early Stopping using the development set, as described in Section~\ref{sec:parameter-estimation}.
Models were trained on a TITAN Xp graphics card.

@@

>>>>

After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction loss on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction loss (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1} = \finitepast\finitefuture$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)},
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $\finitepast$ as in (\ref{eq:z-gauss}), and prediction loss 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] \approx - \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(\finitefuture | Z).
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction loss achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

\subsubsection{Estimating Predictiveness}

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$ \rljf{How?}. Because we consider stationary processes, we have 
that the cross-entropy under $P_\eta$ of $\finitefuture$ is equal to the cross-entropy of $\finitepast$ under the same encoding distribution: $\mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture)\right] = - \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$. \rljf{I see that this is true if $P_\eta$ is the true marginal but what if it is approximate? Could $P_\eta$ end up non-stationary even when it comes from an encoder fit to a stationary distribution? Or is the stationarity of $P_\eta$ guaranteed because the encoder LSTM is autoregressive?}
Using this observation, we estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~\cite{mcallester2018formal}:
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture) - \log P_\psi(\finitefuture | Z)\right],
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

During optimization, we add the cross-entropy term $- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$ to the PRD objective~(\ref{eq:bound}) \rljf{Why?}, so that the full training objective comes out to:
\begin{equation}\label{eq:bound-augmented}
	\min_{\phi, \psi, q,\eta} \left[\mathbb{E}_{\finitepast, \finitefuture} \left[ -	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} \left[ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] - \log P_\eta(\finitepast) \right]\right].
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
Again by Gibbs' inequality and Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}, this is minimized when $P_\eta$ represents the true distribution over length-$M$ blocks $P(\finitepast)$, $P_\phi(Z|\finitepast)$ describes an optimal code for the given $\lambda$, $q$ is its marginal distribution, and $P_\psi(\finitefuture|Z)$ is the Bayes-optimal decoder.
For approximate solutions to this augmented objective, the inequalities~(\ref{eq:bound1-new}-\ref{eq:bound2-new}) will also remain true due to Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}.

@@

Training used Early Stopping using the development set, as described in Section~\ref{sec:parameter-estimation}.
Models were trained on a TITAN Xp graphics card.
\mhahn{TODO number of epochs, training time}

@@

\section{Alternative Modeling Choices}\label{sec:app-choices}

In this section, we investigate alternative modeling choices.

First, we investigated the performance of a simple fixed approximation to the marginal $q$.
We considered a diagonal unit-variance Gaussian, as is common in the literature on Variational Autoencoders~\citep{kingma-auto-encoding-2014}.
We show results in Figure~\ref{fig:even-noFlow}.
Even with this fixed approximation to $q$, NPRD continues to provides estimates not far away from the analytical curve.
However, comparison with the results obtained from full NPRD (Figure~\ref{fig:even}) shows that a flexible parameterized approximation still provides considerably better fit.

Second, we investigated whether the use of recurrent neural networks is necessary.
As recurrent models such as LSTMs process input sequentially, they cannot be fully parallelized, posing the question whether they can be replaced by models that can be parallelized.
Specifically, we considered Quasi-Recurrent Neural Networks, which combine convolutions with a weak form of recurrence, and which have shown strong results on language modeling benchmarks~\citep{bradbury2017quasi}.
We replaced the LSTM with a Quasi-Recurrent Neural Network.
We found that NPRD consistently fitted codes with zero rate when applying this method to the Even Process and the Random Insertion Process.
We also found that the cross-entropies were considerably worse than when using LSTMs.
We conjecture that, due to the convolutional nature of Quasi-Recurrent Neural Networks, they cannot model such processes effectively.

Third, we varied $M$. TODO

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{code/figures/even-info-noFlow.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/rip-info-noFlow.pdf}

%\includegraphics[width=0.45\textwidth]{code/figures/even-beta-mem.pdf}
	\caption{Rate-Distortion for the Even Process (left) and the Random Insertion Process (right), estimated using a simple diagonal unit Gaussian approximation for $q$. Gray lines: Analytical curves. Red dots: Multiple runs of NPRD. Red line: Tradeoff curve computed from NPRD runs. Compare Figure~\ref{fig:even} for results from full NPRD. }\label{fig:even-noFlow}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{code/figures/even-info-varyM.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/rip-info-varyM.pdf}

%\includegraphics[width=0.45\textwidth]{code/figures/even-beta-mem.pdf}
	\caption{Rate-Distortion for the Even Process (left) and the Random Insertion Process (right), varying $M=5, 10, 15$. Gray lines: Analytical curves. Red dots: Multiple runs of NPRD. Red line: Tradeoff curve computed from NPRD runs. Compare Figure~\ref{fig:even} for results from full NPRD. }\label{fig:even-varyM}
\end{figure*}
